The Hydrogen Epoch of Reionization Array (HERA)

### 2015-11-04 manuscript.tex
update to skeleton

\section{Abstract}
\label{sec:abstract}
The Hydrogen Epoch of Reionization Array (HERA http://reionization.org) is a staged experiment that uses the unique properties of the 21-cm line from neutral hydrogen to probe the Epoch of Reionization (EOR). During this epoch, roughly 0.3 - 1 billion years after the Big Bang, the first galaxies and black holes heated and reionized the early Universe. Direct observation of the large scale structure of reionization and its evolution with time will have a profound impact on our understanding of the birth of the first galaxies and black holes, their influence on the intergalactic medium (IGM), and cosmology.  This paper will provide an overview of the project and describe the design of the HERA receiving element.

@@

This intervening period has essentially four phases which may be called the Dark Ages, the Cosmic Dawn, the Epoch of Reionization (EOR) and the Rise of the Galaxies.  The time of the Cosmic Dawn saw the first large-scale gravitational collapse of structures, which eventually heated up the neutral intergalactic hydrogen gas until it again ionized.  This phase change is called the Epoch of Reionization and it represents the last phase change of the Universe as a body.  We may observe this epoch over its extent by observing the red-shifted hydrogen gas.  In our analogy, this is akin to following our child's activities and growth through a good part of their formative years.  Fig \ref{fig:cosmos}.

\begin{figure}[t]
\includegraphics[width=\textwidth]{plots/herauni/herauniall.png}
\caption{History of the Universe  CREDIT BACKGROUND IMAGE - GET PERMISSION
\label{fig:cosmos}}
\end{figure}

@@

HERA has been funded under the US National Science Foundation's Mid-Scale Innovations Program to begin building elements optimized for robust power spectrum detection.  The key is to produce inexpensive collecting area optimized for the spatial scales needed to detect the EOR.  This has resulted in a close-packed array of transit 14-meter dishes (Fig. \ref{fig:array}), with the first elements currently under construction at the South African Karoo Astronomy Reserve, the current location of the Donald C. Backer Precision Array to Probe the Epoch of Reionization (PAPER) array.

Beginning with the current NSF funding, HERA will rollout in a staged deployment.  The first phase extending to August 2016 will deploy 37 elements that will test the HERA element design for performance and manufacturability at location.  These 37 elements will more than double the sensitivity of the 128-element PAPER array.  The next phase extending into 2018 will build out to 127 elements, which should provide a very robust detection of the EOR signal.  Finally, extending into 2020, HERA will build out to 352 elements to further EOR science as a function of redshift and spatial scale, potentially producing the first images of the EOR.

@@

Initial measurements of the EOR strive to measure a statistical power spectrum of the signal since the nature of the reionization process should have a specific spatial signature.  The goal is therefore to measure a range of aggregate spatial scales on the sky, rather than to image the signal directly.  Imaging does remain an ultimate goal to fully understand the process, however we will need a greater understanding of the full measurements to achieve this more difficult goal.  There are now numerous papers in the literature the technique of measuring the power spectrum using an interferometer and the attendant issues (list of citations).  Below, a high-level overview based on [parsons12b] will be presented.

@@

\subsection{Delay-Spectrum Approach}
\label{sec:delayapproach}
Given that the response of an interferometer natively measures the power in Fourier modes of the sky within its beam, we see that it is a natural instrument to use for this measurement.  The so-called delay-spectrum approach leverages this response to optimize sensitivity to the desired modes while rejecting modes contaminated by the foreground power in the wedge.  Other than handling overlapping bins in {\em uv} space, the delay-spectrum approach does not combine baselines before squaring and calculating the power spectrum.

We can essentially relate the sky power spectrum mode ${\hat P}(\kvec)$ to an interferometer baseline visibility ${\tilde V}(\bvec)$, which is the equivalent of the (complex) fringe pattern in a double slit experiment (see {\em e.g.} \cite{2012ApJ...756..165P}):
\begin{equation}
{\hat P}(\kvec) \approx\frac{X^2Y}{4k_B^2}   \left[\frac{{\tilde V}^2(\bvec)}{\Omega_b B/ \lambda^4} \right]
\label{eq:Pk_baseline}
\end{equation}
where $X$ and $Y$ are cosmological parameters relating angular size and spectral frequency to cosmic volumes, $\Omega_b$ is the integrated beam response, $B$ is the effective bandwidth, $\lambda$ is the observation wavelength, and $k_B$ is Boltzmann's constant.  The terms in square brackets are instrumental terms, as opposed to the constants and cosmological parameters.  Rather than ${\hat P}(\kvec)$, the literature generally works with a dimensionless parameter given by $\Delta^2(k) = \frac{k^3}{2\pi^2}{\hat P}(\kvec)$.

@@

\section{Specifications}
\label{sec:specs}
As the initial detection and analysis approach is primarily focused on the delay-spectrum approach, the requirements are set primarily by that approach.  However, since the map-making approaches remain a goal, the design makes every effort to allow such studies.  The high-level specifications are summarized in Fig. \ref{fig:array}.  Specifications fall within one of three categories:  Phase 1, Phase 2, or Extended.

@@

\subsection{Scientific Requirements and Specifications}
\subsubsection{Spectral Response}
Figure \ref{fig:cosmos} highlights the frequency range requirement to probe the expected timescale of the epoch of reionization using the 21cm line of hydrogen as the probe.  
These limits are derived from to-date complementary probes of reionization which include measurements of
the optical depth to last scattering in the CMB, QSO spectra, Ly-$\alpha$
absorption in the spectra of quasars and gamma ray bursts and the demographics
of Ly-$\alpha$ emitting galaxies
%We show limits established by some of these techniques in
(Figure~\ref{fig:IonHist}). 
Constraints from these probes are still weak:
Ly-$\alpha$ absorption saturates at very small neutral fractions; galaxy
surveys directly constrain only the bright end of the luminosity function and
depend on an unknown escape fraction of ionizing photons to constrain
reionization; CMB measurements probe an integral quantity subject
to large degeneracies. Even when these observations are
combined into a single 95\% confidence region, the bounds remain weak.
For example, $x_{HI}$ spans almost the entire allowable range of [0,1]
at $z=8$. 21\,cm reionization experiments place much tighter constraints on ionization, with the
red band showing the forecasted 95\% confidence region derived from HERA data,
after marginalizing over astrophysical and cosmological parameters.

@@

\subsubsection{Spatial/Delay Response}
The spatial response provides the defining characteristics for the experiment; related to the diameter, focal length, baseline, and configuration.  The principal constraint sets the delay response of the antenna element, characterized by the magnitude and frequency of the antenna ripple response.  This was initially addressed to first order in [MEMO] and in a much more detailed way in [UPCOMING PAPERS].  Expressed simply, the delay response of the antenna must be sufficiently attenuated such that the foreground signal levels at those delays is below the expected signal strength of the EOR signal.  This sets the optics and match needs.  As discussed in [REFS], this is addressed by direct measurement and modeling of the response to the system relative to and as a function of delay time from the initial wave reception.  The expected strength of the foregrounds is modeled through the system and plotted on the same set of axes.  These two regions should be disjoint to a significance of 60 dB.

Spatial modes - delay 60@60, baseline, diameter

\subsubsection{Sensitivity}
sensitivity - reionization fraction 10$\sigma$ from 0.2-0.8

@@

\subsection{Programmatic Specifications}
stages/soon - phase with initial PAPER equipment

cost - cheap experiment

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=7.5cm]{plots/costfigNew.png} 
}
\caption{\small Costing model for a fixed sensitivity system as a function of diameter (thick blue line and left axis) as well as the round-trip delay for 1,2 and 3 feed-vertex trips (dashed red lines and right axis).  The 60 ns spec is indicated by the horizontal dash line, and the two-trip intersection at about 14-meter is the vertical dashed line.
\label{fig:costfig}}
\end{figure}

@@

>>>>

\section{Abstract}
\label{sec:abstract}
The Hydrogen Epoch of Reionization Array (HERA http://reionization.org) is a staged experiment that uses the unique properties of the 21-cm line from neutral hydrogen to probe the Epoch of Reionization (EOR). During this epoch, roughly 0.3 - 1 billion years after the Big Bang, the first galaxies and black holes heated and reionized the early Universe. Direct observation of the large scale structure of reionization and its evolution with time will have a profound impact on our understanding of the birth of the first galaxies and black holes, their influence on the intergalactic medium (IGM), and cosmology.  This paper will provide an overview of the project and describe the design of the system.

@@

This intervening period has essentially four phases which may be called the Dark Ages, the Cosmic Dawn, the Epoch of Reionization (EOR) and the Rise of the Galaxies.  The time of the Cosmic Dawn saw the first large-scale gravitational collapse of structures, which eventually heated up the neutral intergalactic hydrogen gas until it again ionized.  This phase change is called the Epoch of Reionization and it represents the last phase change of the Universe as a body.  We may observe this epoch over its extent by observing the red-shifted hydrogen gas.  In our analogy, this is akin to following our child's activities and growth through a good part of their formative years.  Fig. \ref{fig:cosmos} shows a depiction of the history of the Universe, as well as the 21-cm Hydrogen line red-shifted to cosmic time.

\begin{figure}[t]
\includegraphics[width=\textwidth]{plots/herauni/herauniall.png}
\caption{history of the Universe plotted on a log scale.  The top axis shows the time elapsed since the Big Bang while the bottom shows the corresponding redshift.  Of course, placing exactly when the depicted events occurred and how is the goal of HERA.  The right axis and the red-line show the frequency of the 121-cm Hydrogen line over cosmic time.  The horizontal lines denote the frequency range of HERA, both the initial band (the inner pair) and the desired extended band.  CREDIT BACKGROUND IMAGE - GET PERMISSION
\label{fig:cosmos}}
\end{figure}

@@

HERA has been funded under the US National Science Foundation's Mid-Scale Innovations Program to begin building elements optimized for robust power spectrum detection.  The key is to produce inexpensive collecting area optimized to detect the EOR.  This has resulted in a close-packed array of transit 14-meter dishes (Fig. \ref{fig:array}), with the first elements currently under construction at the South African Karoo Astronomy Reserve, the current location of the Donald C. Backer Precision Array to Probe the Epoch of Reionization (PAPER) array.  [PAPER REF]

Beginning with the current NSF funding, HERA is rolling out in a staged deployment.  The first phase extending to August 2016 will deploy 37 elements that will test the HERA element design for performance and manufacturability at location.  These 37 elements will more than double the sensitivity of the 128-element PAPER array.  The next phase extending into 2018 will build out to 127 elements, which should provide a very robust detection of the EOR signal.  Finally, extending into 2020, HERA will build out to 352 elements to further EOR science as a function of redshift and spatial scale, potentially producing the first images of the EOR.

@@

Given the difficulty of the experiment, initial measurements of the EOR strive to measure a statistical power spectrum of the signal since the nature of the reionization process should have a specific spatial signature.  The goal is therefore to measure a range of aggregate spatial scales on the sky, rather than to image the signal directly.  Imaging does remain an ultimate goal to fully understand the process, however we will need a greater understanding of the full measurements to achieve this more difficult goal.

Obviously between our present observation point and the Epoch of Reionization approximately 13 billion years ago lies the entire intervening Universe, which has an intrinsically much brighter signal (up to 6 orders of magnitude or more).  Primarily, this power is due to diffuse Galactic synchrotron radiation, supernova remnants and extragalactic radio sources.  As a first step, areas of the sky with where these signals are minimal (for example, outside the galactic plane and away from strong point sources) are targeted (REF FOR THIS).  More importantly and most fortuitously however is the fact that all of these signals are smooth spectrum sources whereas the expected spectrum of the EOR is expected to be rough since it is made up of non-ionized regions which are randomly distributed over a wide range of redshifts.  This fact allows us to try and isolate foregrounds from the EOR, as will be discussed.

@@

\subsection{Delay-Spectrum Approach}
\label{sec:delayapproach}
Given that the response of an interferometer natively measures the power in Fourier modes of the sky within its beam, we see that it is a natural instrument to use for this measurement.  There are now numerous papers in the literature the technique of measuring the power spectrum using an interferometer and the attendant issues (list of citations).  Below, a high-level overview based on  \cite{2012ApJ...756..165P} will be presented. The so-called delay-spectrum approach leverages this response to optimize sensitivity to the desired modes while rejecting modes contaminated by the foreground power in the wedge.  Other than handling overlapping bins in {\em uv} space, the delay-spectrum approach does not combine baselines before squaring and calculating the power spectrum.

We can essentially relate the sky power spectrum mode ${\hat P}(\kvec)$ to an interferometer baseline visibility ${\tilde V}(\bvec)$, which is the equivalent of the (complex) fringe pattern in a double slit experiment:
\begin{equation}
{\hat P}(\kvec) \approx\frac{X^2Y}{4k_B^2}   \left[\frac{{\tilde V}^2(\bvec)}{\Omega_b B/ \lambda^4} \right]
\label{eq:Pk_baseline}
\end{equation}
where $X$ and $Y$ are cosmological parameters relating angular size and spectral frequency to cosmic volumes, $\Omega_b$ is the integrated beam response, $B$ is the effective bandwidth, $\lambda$ is the observation wavelength, and $k_B$ is Boltzmann's constant.  The terms in square brackets are instrumental terms, as opposed to the constants and cosmological parameters.  Rather than ${\hat P}(\kvec)$, the literature generally works with a volume-normalized parameter given by $\Delta^2(k) = \frac{k^3}{2\pi^2}{\hat P}(\kvec)$.

@@

\section{Scientific Requirements and Specifications}
\label{sec:specs}
As the initial detection and analysis approach is primarily focused on the delay-spectrum approach, the requirements are set primarily by that approach.  However, since the map-making approaches remain a goal, the design makes every effort to allow such studies.  The high-level specifications are summarized in Fig. \ref{fig:array}.  Specifications fall within one of three categories:  Phase 1, Phase 2, or Extended.

@@

\subsection{Spectral Response}
Figure \ref{fig:cosmos} highlights the frequency range requirement to probe the expected timescale of the epoch of reionization using the 21cm line of hydrogen as the probe.  
These limits are derived from to-date complementary probes of reionization which include measurements of
the optical depth to last scattering in the CMB, QSO spectra, Ly-$\alpha$
absorption in the spectra of quasars and gamma ray bursts and the demographics
of Ly-$\alpha$ emitting galaxies
%We show limits established by some of these techniques in
(Figure~\ref{fig:IonHist}). 
Constraints from these probes are still weak:
Ly-$\alpha$ absorption saturates at very small neutral fractions; galaxy
surveys directly constrain only the bright end of the luminosity function and
depend on an unknown escape fraction of ionizing photons to constrain
reionization; CMB measurements probe an integral quantity subject
to large degeneracies. Even when these observations are
combined into a single 95\% confidence region, the bounds remain weak.
For example, $x_{HI}$ spans almost the entire allowable range of [0,1]
at $z=8$. 21\,cm reionization experiments place much tighter constraints on ionization, with the
red band showing the forecasted 95\% confidence region derived from HERA data,
after marginalizing over astrophysical and cosmological parameters.

@@

\subsection{Spatial/Delay Response}
The spatial response provides the defining characteristics for the experiment; related to the diameter, focal length, baseline, and configuration.  The principal constraint sets the delay response of the antenna element, characterized by the magnitude and frequency of the antenna ripple response.  This was initially addressed to first order in [MEMO] and in a much more detailed way in [UPCOMING PAPERS].  Expressed simply, the delay response of the antenna must be sufficiently attenuated such that the foreground signal levels at those delays is below the expected signal strength of the EOR signal.  This sets the optics and match needs.  As discussed in [REFS], this is addressed by direct measurement and modeling of the response to the system relative to and as a function of delay time from the initial wave reception.  The expected strength of the foregrounds is modeled through the system and plotted on the same set of axes.  These two regions should be disjoint to a significance of 60 dB.

The minimum baseline is set such that the baseline has negligible impact on the minimum baseline (shown in Fig. \ref{fig:kperf} in the standard analysis.  Somewhat arbitrarily, a maximum shortest baseline with equal contribution to $k_{min}$ at the full bandwidth of 100 MHz is chosen, which corresponds to about 15m.  Additionally, this baseline has been well-studied by the PAPER array (ref).

The above requirements also drive the specification on the element diameter, as well as the desire to maximize sensitivity.  This means that the diameter should be something around 15m (that is, a fully filled aperture for the minimum baseline) with a short focal length.

\subsection{Sensitivity}
sensitivity - reionization fraction 10$\sigma$ from 0.2-0.8

@@

\subsection{Cost}
\begin{figure}[t]
\centerline{
\includegraphics[width=7.5cm]{plots/costfigNew.png} 
}
\caption{\small Costing model for a fixed sensitivity system as a function of diameter (thick blue line and left axis) as well as the round-trip delay for 1,2 and 3 feed-vertex trips (dashed red lines and right axis).  The 60 ns spec is indicated by the horizontal dash line, and the two-trip intersection at about 14-meter is the vertical dashed line.
\label{fig:costfig}}
\end{figure}

@@

@@

\subsection{Configuration}
\label{sec:config}
